Labor Markets and Technological Change: Evidence from Electronic Health Records

### 2024-09-07 Paper/PaperEdit.tex

\title{Labor Markets and Technological Change: Evidence from Electronic Health Records}
\author{Hanna Glenn}
%\date{\today}

@@

\begin{abstract}
    Technology adoption in health care is of great importance to policy makers as some technologies are expected to radically improve the U.S. health care system. However, some new technologies can be burdensome for physicians and lead to frustration. I add to our understanding of a new technology's affect on workers by investigating changes in physician behavior as a result of a major technology shift in U.S. hospitals, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes. I find that hospitalists are 10\% more likely to retire, and at least 15\% more likely to work in an office due to EHRs. Further, those exposed to EHRs see more patients due to the technology. These results point to unintended consequences of rapid technology implementation, but also indicate potential productivity gains for those who embrace the technology.
\end{abstract}

@@

EHRs are computer systems that house digital medical records and sometimes have additional functionalities. EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{According to Quatris Healthco, meaningful use standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy. For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. Daily tasks change drastically when a new system is put in place; some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}). When looking solely at hospitalists, 63\% of survey respondents found EHRs time consuming and inefficient (\cite{czernik2022hospitalist}). The frustration of a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. For a hospitalist, the hospital's decision to adopt an EHR is plausibly exogenous. Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical environments, measured based on no future claims, (2) substituting hospital work for office work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.

@@

I expand this literature first by considering individual physician labor market decisions as an outcome, which, to my knowledge, has not yet been examined. Further, I consider the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting physicians' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework. I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this result, I find that EHR implementation led to a 20\% increase in the likelihood of exiting clinical settings for hospitalists of age 60 plus. Physicians less than 60 were 7\% more likely to exit, possibly pointing to career switching. Limiting the sample to those who remained in clinical settings, I find that hospitalists were 30\% more likely to start seeing patients in an office setting at the time of exposure. Finally, I find that EHRs lead to a patient count increase of at least 20 patients per year, and the increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individuals. That is, a hospitalist employed by the hospital does not choose their own EHR exposure. My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of experienced hospitalists who left both clinical settings altogether and the hospitals where the EHR was deployed. Thus, EHR implementation may have intensified access to care issues for patients. This unintended consequence highlights the need for careful consideration of technologies and policies that affect work environment.  \textcolor{red}{reword the implications and check the numbers are correct}

@@

Research on how physicians interact with EHRs is primarily focused on outpatient settings. On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming portions of EHR use (\cite{arndt2017tethered}). In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR. Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use. The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}). Another common issue physicians face is the lack of usability of EHR systems, that functions take minutes to either locate or load. Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult. While most research is specific to physicians in outpatient settings, the limited research available which investigates hospitalists and EHRs seems to be consistent with these findings (\cite{tipping2010did}).

@@

For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting. However, there are two types of physicians who can be categorized as hospitalists based on this definition: one being a formal hospitalist, and the other being an internist. Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings. Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting. While this distinction is potentially important for some of the decisions outlined below, the data do not explicitely distinguish between the two. I will discuss this limitation in more detail when relevant.

@@

Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}, \cite{loria_2020}), but this has not been addressed empirically.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.). I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}). I use this data to construct the dependent variables for my analysis. Generally, these include the decision to exit clinical work, switch to a different work setting, number of patients seen, and billing activity.

@@

I aggregate the data to the physician level. I drop physicians who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated. The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare (provided by CMS). I limit to physicians who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

@@

\normalsize
I follow a maximum of 19,263 hospitalists over time, depending on the sample considered. Out of all hospitalists in the sample, 3\% (approximately 578) stop seeing patients at some point. A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards early years. For hospitalists in the office setting sample (those who do not exit), 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office. There are 12,732 hospitalists who never exit or change their work setting at any point in the sample. These hospitalists see 325 patients and file 4 claims per patient on average. Consistent across all samples, hospitalists tend to work with 1.3 hospitals and 1.2 systems.

@@

I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of exiting clinical work for the full sample of hospitalists, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age in Figure \ref{fig:retirefirst}.\footnote{Alternatively, one could estimate the effect for more age groups and observe the relationship between the overall ATT and age. However, the data does not support further sub-setting as sample sizes become small.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}. Being exposed to an EHR leads to a .2 ppt increase in the likelihood of exiting in the first year after exposure, and a .4 ppt increase in the second year after exposure. While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03. That is, the effect represents a 7\% (13\% in the second year) increase in the likelihood of exiting relative to the mean.

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Retirement}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.55]{Objects/retire_plot_all.pdf}
   \label{fig:retirefirst} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.55]{Objects/retire_plot_ages.pdf}
   \label{fig:retiresecond}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} Panel (a) shows average group time treatment effects aggregated over groups to an event study plot. Panel (b) shows these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0009), 0.002 (0.001), 0.006 (0.003), respectively.}}
\end{figure}

Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$). Interestingly, hospitalists >=60 are driving the positive estimate in the year after exposure, and younger hospitalists only exit in the second year after exposure. A retirement age physician is 1 ppt more likely to exit after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, hospitalists less than 60 are not more likely to exit in the first year after exposure, but are approximately 16\% more likely to exit two years after exposure. There are two potential reasons for the different response time that relate to formal retirement vs. career switching. First, if younger physicians are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year. Similarly, younger hospitalists may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.

There is no visual indication of a violation of the parallel trends assumption in the pre-periods. However, the p-value of a Wald test for pre-trends indicates some evidence for such a violation for younger hospitalists. In Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends. Due to the rarity of exiting for this sub-sample, the estimate becomes null with any specified variation in the parallel trend assumption. I still discuss the estimates for this age group, but focus on the results for the older hospitalists. I also present overall treatment effects under different data specifications in Appendix \ref{sec:chart}, and find that this result is robust to various specifications.

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.55]{Objects/officeind_plot_all.pdf}
   \label{fig:officefirsta} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.55]{Objects/officeind_plot_ages.pdf}
   \label{fig:officefirstb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .07 (0.008), .06 (0.008), 0.12 (0.03), respectively}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.55]{Objects/officefrac_plot_all.pdf}
   \label{fig:officeseconda} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.55]{Objects/officefrac_plot_ages.pdf}
   \label{fig:officesecondb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .08 (0.003), .07 (0.003), 0.13 (0.01), respectively.}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Number of Patients Seen}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.55]{Objects/patient_plot_all.pdf}
   \label{fig:patienta} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.55]{Objects/patient_plot_ages.pdf}
   \label{fig:patientb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 28 (5.0), 25.6 (5.2), 40.4 (15.0), respectively.}}
\end{figure}

@@

Finally, I consider whether EHR exposure affects claims filed per patient. On average, hospitalists file roughly 4 claims per patient. I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time. The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as hospitalists are exposed to the new software, they bill for more items on average.  I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists.

There is no evidence of a pre-trend driving this result, and I also show robustness to violations in parallel trends in Appendix \ref{sec:pretrends}. I also compute \citeauthor{lee2009training} (\citeyear{lee2009training}) bounds for this outcome due to the limited sample. The computed bounds on the average treatment effect are .45 to .6 claims per patient, compared to the overall effect in the main specification of .33. The effect becomes even larger when accounting for a particular type of physician being left in the sample. However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, around half of the estimates are null. Anticipation and limiting sample years cause the effect to disappear or even become negative. Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs.

>>>>

\title{Labor Markets and Technological Change: Evidence from Electronic Health Records}
\author{Hanna Glenn}

@@

\begin{abstract}
    Technology adoption in health care has the potential to improve patient outcomes and, in some cases, decrease costs. However, new technologies can be burdensome for physicians to keep up with and lead to frustration. I add to our understanding of how new technology can affect those who use it by investigating changes in physician behavior as a result of a major technology shift in U.S. hospitals, electronic health records (EHRs). I treat EHR implementation in hospitals as an exogenous treatment to physicians within the hospital (hospitalists), and estimate average group time treatment effects on various labor market outcomes. I find that hospitalists are more likely to retire, and more likely to shift work to an outpatient setting due to EHRs. However, those who continue using EHRs in the hospital see more patients due to the technology. 
\end{abstract}

@@

EHRs are computer systems that house digital medical records and sometimes have additional functionalities. EHRs have become increasingly relevant in the US since 2009, when the Health Information Technology for Economic and Clinical Health (HITECH) Act was passed to subsidize hospitals and practices who implement and ``meaningfully'' use an EHR (\cite{hitech}).\footnote{Meaningful use (MU) standards proceeded in three stages over time. In Stage 1 (2010), MU focused on data capturing and sharing. In Stage 2, which began in late 2012, MU extended to using EHRs for patient incorporation and using the technology as a helper in care. Stage 3 went from 2014-2016 and focused on making data accessible across hospitals (\cite{meanuse})} President Obama stated in 2009, “To improve the quality of our health care while lowering its cost, we will make the immediate investments necessary to ensure that, within five years, all of America’s medical records are computerized.” (\cite{presquote}). The push for EHR use stemmed from a widespread expectation that EHRs would improve quality of health care while decreasing costs, the gold standard in health policy. For example, a 2005 study estimated hundreds of billions of dollars saved if health information technology were to be fully implemented (\cite{hillestad2005}).

Physicians, as the primary users of EHRs, play an important role in whether the expected benefits come to fruition. Daily tasks change drastically when a new system is put in place; some physicians reported that when using EHRs they are less satisfied with their job and have higher stress levels. Senior physicians in particular “loathe the cumbersome, time-consuming data entry that comes with using EHR.” (\cite{CollierBurnout}). When looking solely at hospitalists, 63\% of survey respondents found EHRs time consuming and inefficient (\cite{czernik2022hospitalist}). The frustration of a new technology raises the cost of working in settings that use it, which may lead physicians on the margin to make behavioral changes such as exiting the labor market altogether or shifting towards alternative work settings. However, the extent to which this frustration actually imposes a meaningful cost is unknown, making physician response an empirical question. For a hospitalist, the hospital's decision to adopt an EHR is plausibly exogenous. Thus, I use a difference-in-differences research design to estimate the effect of EHR adoption on hospitalist labor market choices: (1) exiting clinical work altogether, measured based on no future claims, (2) substituting hospital work for office/outpatient work, measured by the location of claims, and (3) productivity, measured by the number of patients seen and claims filed per patient.

@@

I expand this literature first by considering individual physician labor market decisions as an outcome, which, to my knowledge, has not yet been examined. Further, I consider the primary time period in which EHRs were rapidly implemented. This provides sufficient variation in the timing of adoption and ensures the EHRs are meaningful and are likely affecting hospitalists' daily life. Similarly to past studies, I consider EHR implementation as a treatment variable in a difference-in-differences framework. I improve on this strategy by estimating average group time treatment effects for multiple years after implementation, which avoids common problems that arise in typical two way fixed effects estimation with heterogeneous effects.

Early research on the inputs to physician retirement decisions found that the most influential factors are personal and financial matters, and that physicians care about their work environment (specifically in the context of managed care organizations), but do not necessarily retire early because of it (\cite{Bahrami2002}). Contrary to this result, I find that EHR implementation led to a 20\% increase in the likelihood of exiting clinical settings for hospitalists of age 60 plus. Physicians less than 60 were 7\% more likely to exit, possibly pointing to career switching. Limiting the sample to those who remained in clinical settings, I find that hospitalists were 30\% more likely to start seeing patients in an office setting at the time of exposure. Finally, I examine productivity of hospitalists who do not exit or change work settings, and I find that EHRs lead to a patient count increase of at least 20 patients per year. This increase is large enough that it is not entirely driven by a redistribution of patients from those who left clinical settings.

Two key assumptions underlie the main analysis. First, I assume that hospitalists are constrained to use the technology implemented in their work setting. Second, I assume the decision for hospitals to implement EHRs is exogenous to individual hospitalists. That is, a hospitalist employed by the hospital does not choose their own EHR exposure. My main findings are not sensitive to a plethora of robustness checks, and the results suggest that there was a surge of hospitalists who left clinical settings altogether, and even more that shifted work away from the hospitals where the EHR was deployed and towards outpatient/office settings. Thus, EHR implementation may have intensified access to care issues for hospitalized patients. This unintended consequence of incentivizing rapid technology adoption highlights the need for careful consideration of technologies and policies that might negatively affect work environment.

@@

Research on how physicians interact with EHRs is primarily focused on outpatient settings. On a daily basis, a physician spends approximately 23.7\%, 17\%, and 15.5\% of their time on documentation, chart review, and inbox management, respectively, which are the most time consuming portions of EHR use (\cite{arndt2017tethered}). In all, if a physician works for 12 hours, they spend roughly 3.2 hours with patients and 5.9 hours interfacing with an EHR. Despite a large amount of time spent with them, physicians still continue to report frustration over EHR use. The most time consuming functions are also reported as the most frustrating (\cite{dymek2021building}). Another common issue physicians face is the lack of usability of EHR systems, with functions taking minutes to locate or load. Hospitals often provide training upon adoption, but rapid system upgrades and changes make familiarity with the system difficult. While most research is specific to physicians in outpatient settings, the limited research available which investigates hospitalists and EHRs seems to be consistent with these findings (\cite{tipping2010did}).

@@

For ease of notation, I define a hospitalist in my data as any general practice physician who sees at least 70\% of their patients within a hospital setting. However, there are two types of physicians who can be categorized as hospitalists based on this definition: a formal hospitalist who is solely employed by the hospital, or an internist who contracts their services to the hospital. Both formal hospitalists and internists are medical professionals who primarily focus on general patient care, but they potentially differ in their work settings. Internists are primary care physicians who provide comprehensive medical care to adult patients in either outpatient settings or hospitals, whereas formal hospitalists specialize in the care of hospitalized patients, and typically work exclusively in the hospital setting. While this distinction is potentially important for some of the decisions outlined below, the data do not explicitly distinguish between the two. I will discuss this limitation in more detail when relevant.

@@

Age is the most important factor in the decision to retire, implying that a 35 year old physician has very different incentives than a 65 year old physician, and a technology shock will affect the age groups differently. As a worker considers the retirement decision, they weigh their expected earnings with their expected costs, along with some preferences. Then, a physician will only retire if the utility from working is less than the utility from retiring, where utility partially depends on job satisfaction and personal life factors. Exogenous EHR implementation affects the decision to retire through job satisfaction. If EHRs impose a burden on daily tasks, there is an inverse relationship between EHR exposure and utility from continuing to work. If EHRs make daily tasks easier, there is a positive relationship. Thus, for a physician indifferent between retiring and continuing to work, EHRs induce retirement if they impose a cost in terms of job satisfaction, or induce continued working life if they impose a benefit to job satisfaction. Media articles focusing on interviews with physicians suggest that the implementation of EHRs did in fact lead to retirement in older physicians (\cite{ringel_2019}; \cite{loria_2020}), but this has not been addressed empirically.

@@

The Medicare Data on Provider Practice and Specialty (MD-PPAS) is a database of physicians which details claim counts, physician specialty, and practice location from 2009-2017. Variables included are physician NPI, primary and secondary specialties, patient count and total billing counts, and fraction of patients seen in different settings (hospital, office/outpatient, emergency room, etc.). I only include physicians classified as hospitalists\footnote{Physician who self-identifies as hospitalist or has 90\% of patients in inpatient setting. I cannot distinguish between types of hospitalists as the data source manipulates this definition based on percent of patients seen in a hospital, so these can be formal hospitalists or internists.}, internal medicine, general practice or family practice in at least one year. I drop physicians with less than 70\% of patients in a hospital setting in all years, as the focus is on those who are affected by a hospital's decision to adopt (I investigate this threshold in Appendix \ref{sec:changes}). I use this data to construct the dependent variables for my analysis.

@@

I aggregate the data to the hospitalist level. I drop those who are not treated by 2015 since I do not observe whether they become treated in 2016 or 2017.\footnote{In Appendix \ref{sec:changes} I limit years to 2009-2015 and include never-treated. The results are robust to this specification choice.} That is, the data contains no units which are ``never-treated". Finally, I include a physician's graduation year from Physician Compare (provided by CMS). I limit to hospitalists who graduated before 2005, as anyone who graduated medical school after that will be finishing residency during the span of the data and will exhibit labor market changing behavior which could be correlated to EHR exposure purely because of switching work settings during a time of rapid EHR implementation.There are various thresholds used in the construction of this final data, including how many shared patients constitutes a physician-hospital relationship. The details of such thresholds are explained in Appendix \ref{app:data}. To ensure that my results  are not sensitive to various levels of the thresholds, I estimate the effect of EHR exposure on all outcomes under different combinations of data limitations. The specification charts with these results are shown in Appendix \ref{sec:chart}, and I discuss each in conjunction with the main results.

@@

\normalsize
I follow a maximum of 19,263 hospitalists over time, depending on the sample considered. Out of all hospitalists in the sample, 3\% (approximately 578) stop seeing patients at some point. A positive number of hospitalists are exposed to an EHR in each year, but the majority of exposure is concentrated towards early years. For hospitalists in the office setting sample (those who do not exit), 30\% of physicians have a positive number of patients seen in an office, and the average physician sees 9\% of their total patient volume in an office. There are 12,732 hospitalists who never exit or change their work setting at any point in the sample. These hospitalists see 325 patients and file 4 claims per patient on average. Consistent across all samples, hospitalists tend to work approximately 1.5 hospitals and 1.3 systems.

@@

I present aggregated group time treatment effects of being exposed to an EHR on the likelihood of exiting clinical work for the full sample of hospitalists, as well as split samples for those at least 60 (typical retirement age) and less than 60 years of age in Figure \ref{fig:retirefirst}.\footnote{Alternatively, one could estimate the effect for more age groups and observe the relationship between the overall ATT and age. However, the data does not support further sub-setting as sample sizes become small.} These effects are a weighted average of the $ATT(g,t)$ parameters discussed in Section \ref{sec:empstrat}. Being exposed to an EHR leads to a .2 ppt increase in the likelihood of exiting in the first year after exposure, and a .4 ppt increase in the second year after exposure. Coefficients remain similar 3 and 4 years after exposure, though estimates become noisier due to smaller sample not-yet-treated groups.13 While numerically small, these effects are economically meaningful relative to the proportion of hospitalists who exit in the sample, .03. That is, the effect represents a 7\% (13\% in the second year) increase in the likelihood of exiting relative to the mean.

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on No Longer Seeing Patients}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.5]{Objects/retire_plot_all.pdf}
   \label{fig:retirefirst} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.5]{Objects/retire_plot_ages.pdf}
   \label{fig:retiresecond}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} Panel (a) shows average group time treatment effects aggregated over groups to an event study plot. Panel (b) shows these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 0.003 (0.0009), 0.002 (0.001), 0.006 (0.003), respectively.}}
\end{figure}

Next, I examine how the results differ when considering physicians in different age brackets: pre-retirement ($<= 60$) and typical retirement age ($> 60$). Interestingly, hospitalists $>=60$ are driving the positive estimate in the year after exposure, and younger hospitalists only exit in the second year after exposure. A retirement age hospitalist is 1 ppt more likely to exit after being exposed to an EHR, a 25\% increase relative to the mean. In comparison, hospitalists less than 60 are not more likely to exit in the first year after exposure, but are approximately 16\% more likely to exit two years after exposure. There are two potential reasons for the different response time that relate to formal retirement vs. career switching. First, if younger hospitalists are leaving clinical work for administrative or consulting roles, the preparation time for this type of switch (preparing a resume, applying for new jobs, etc.) could be longer than one year, whereas formal retirement can be decided and executed within one year. Similarly, younger hospitalists may attempt to learn complex EHRs for a longer period of time before deciding the cost is too high.

There is no visual indication of a violation of the parallel trends assumption in the pre-periods. However, the p-value of a Wald test for pre-trends indicates some evidence for such a violation for younger hospitalists. In Appendix \ref{sec:pretrends}, I present confidence intervals robust to various specifications of a violation in parallel trends. Due to the rarity of exiting for this sub-sample, the estimate becomes null with any specified variation in the parallel trend assumption. I still discuss the estimates for this age group, but focus on the results for the older hospitalists. I also present overall treatment effects under different data specifications in Appendix \ref{sec:chart}. Under various data and specification changes, the coefficients remain non-negative. Estimates become noisier under strict limitations of the data, but given the rarity of exiting, this positive effect still seems to be robust.

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Likelihood of Working in Office}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.5]{Objects/officeind_plot_all.pdf}
   \label{fig:officefirsta} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.5]{Objects/officeind_plot_ages.pdf}
   \label{fig:officefirstb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .07 (0.008), .06 (0.008), 0.12 (0.03), respectively}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Fraction of Patients Seen in Office}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.5]{Objects/officefrac_plot_all.pdf}
   \label{fig:officeseconda} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.5]{Objects/officefrac_plot_ages.pdf}
   \label{fig:officesecondb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: .08 (0.003), .07 (0.003), 0.13 (0.01), respectively.}}
\end{figure}

@@

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Number of Patients Seen}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.5]{Objects/patient_plot_all.pdf}
   \label{fig:patienta} 
\end{subfigure}

@@

\begin{subfigure}[]{.8\textwidth}
\caption{sub-samples of physicians based on age}
   \includegraphics[scale=.5]{Objects/patient_plot_ages.pdf}
   \label{fig:patientb}
\end{subfigure}
\vspace{2mm}
    \caption*{\footnotesize{\textit{Notes:} The top panel shows average group time treatment effects aggregated over groups to an event study plot. The bottom show these results for different subgroups of physicians by age. The p-value listed for each graph corresponds to a Wald test for pre-trends. Confidence intervals shown are simultaneous confidence bands accounting for multiple hypothesis testing. Overall ATT for all, $<$ 60, $>=$ 60 with SE in parentheses: 28 (5.0), 25.6 (5.2), 40.4 (15.0), respectively.}}
\end{figure}

@@

Finally, I consider whether EHR exposure affects claims filed per patient. On average, hospitalists file roughly 4 claims per patient. I present estimates in Figure \ref{fig:claima}, which shows that EHRs tend to increase claims filed per patient over time. The magnitude of the effect ranges from .2 to .8, or 5-20\% relative to the mean, indicating that as hospitalists are exposed to the new software, they bill for more items on average.  I also show heterogeneity of this result by age group, and Figure \ref{fig:claimb} shows that younger hospitalists are exhibiting more of an increase in claim count than older hospitalists. However, when I analyze this relationship under different data specifications (shown in Appendix \ref{sec:chart}, a majority of the estimates are negative or zero. Anticipation and limiting sample years cause the effect to disappear or even become negative. Thus, it is not clear whether this positive effect is actually an unintended consequence of EHRs and I do not make any strong conclusions.

\begin{figure}[ht!]
\centering
    \captionsetup{width=.85\linewidth}
    \caption{Effect of EHR Exposure on Claims per Patient}
   \begin{subfigure}[]{.8\textwidth}
   \caption{all physicians}
   \includegraphics[scale=.5]{Objects/claim_per_patient_plot_all.pdf}
   \label{fig:claima} 
\end{subfigure}
